---
title: "Wide Open Spaces"
output: pdf_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```



```{r , echo=FALSE, include = FALSE}
library(ggplot2)
library(tidyverse)
library(ggmap)
library(leaflet)
library(zipcodeR)
library(tidycensus)
library(foreach)
library(stringr)
library(dplyr)
library(NbClust)
library(cluster)
library(kableExtra)
library(formattable)
```


```{r, echo=FALSE, include=FALSE}
#Google API key
ggmap::register_google(key = "AIzaSyDjpa_BSzUs11JA1rBa2YFZs3YMrfPiu-A")

#imports data for Austin parks and median income per zipcode
austin_parks = read.csv('/Users/franklinstudent/Desktop/GitHub/Parks_Project/City_of_Austin_Parks_data.csv')
travis_parks = read.csv('/Users/franklinstudent/Desktop/GitHub/Parks_Project/Travis_Parks.csv')
more_parks = read.csv('/Users/franklinstudent/Desktop/GitHub/Parks_Project/more_parks.csv')
austin_medianincome = read.csv('/Users/franklinstudent/Desktop/GitHub/Parks_Project/Austin_median_income.csv')
```

# I. Abstract
The objective of this paper is to discern a correlation between post-secondary education and wealth, and equal opportunity to access outdoor spaces in Austin, Texas. Obtaining data derived from the parks departments of several localities and the Austin Business Journal, hierarchical clustering was used to determine if there existed an imbalance between areas of Austin. Information of various characteristics within a City of Austin zip code were important tools in analyzing the data. In aggregate, there exists 335 outdoor spaces spread across 43 zip codes.

# II. Introduction
For more than a year, Covid-19 has created a environment in which many around the world remained in isolation from one another. In an effort to find respite, many have ventured to outdoor spaces, but for others, these spaces may be less easily experienced. The objective of this paper is to determine whether Austinites have equal opportunity to access outdoor spaces. To accomplish this objective, I used data derived from the parks departments for the City of Austin and Travis County. Moreover, the Austin Business Journal provided 2018 data specific to each zip code of relevant characteristics, such as educational attainment, median household income, and median home value.

# III. Methods
Since individuals seek relaxation in a wide variety of ways, I applied a wide definition in determining an outdoor space. Within Austin, these not only include local and state parks, but also include public pools and private golf courses - totaling 335 outdoor spaces. Of these locations, 319 were obtained through the parks departments for the City of Austin and Travis County. However, the remaining 16 locations were derived individually from Google Maps and corresponding parks departments of Bee Cave and Wells Branch. Although, Bee Cave and Wells Branch are not technically within the city limits of Austin, their respective zip codes are categorized as city of Austin zip codes, thus both Bee Cave and Wells Branch were included in the data set.

The 43 zip codes for the City of Austin became a useful tool in the methodology applied to the model. This information was derived from the Austin Business Journal, which provided data of relevant characteristics, such as educational attainment, median household income, and median home value. In addition to using several of these features and data of outdoor spaces, I created an interactive variable between post-secondary educational attainment as a percentage of zip code population and median household income, mapping the outcome on Google Maps so as to discover any patterns.

To best solve the objective of this paper, I determined hierarchical clustering was the best approach. First, I created a large data set of numerical vectors, in which I then centered and scaled before using the euclidean method to find the distances between each point. Next, I determined the Ward method was appropriate in use of hierarchical clustering as it produced a well-balanced cluster denodrogram. Lastly, I used k-means and the elbow method to find the optimal k to cut the tree, and discerned k = 2 would be optimal. 



```{r, echo=FALSE}
#removes unnecessary columns
austin_parks = select(austin_parks, -(1:2), -(4:29), -(31:45), -(47:49), -(51:55))
travis_parks = select(travis_parks, -1, -3, -(6:20))
more_parks = select(more_parks, -(4:5))
austin_medianincome = select(austin_medianincome,-(4:5))
```



```{r, echo=FALSE}
#renames column
austin_medianincome = rename(austin_medianincome, zipcode = ZIP.Code)
austin_medianincome = rename(austin_medianincome, Population = Population..2018.Est..1)

#converts variables to numeric
austin_medianincome$Median.Household.Income <- as.numeric(gsub('[$,]', '', austin_medianincome$Median.Household.Income))
austin_medianincome$Population <- as.numeric(gsub(',', '', austin_medianincome$Population))
austin_medianincome$No..Of.Households..2018.Est.<- as.numeric(gsub(',', '', austin_medianincome$No..Of.Households..2018.Est.))
austin_medianincome$Per.Capita.Income..2018.Est. <- as.numeric(gsub('[$,]', '', austin_medianincome$Per.Capita.Income..2018.Est.))
austin_medianincome$Median.Value.Of.Owner.Occupied.Homes..2018.Est. <- as.numeric(gsub('[$,]', '', austin_medianincome$Median.Value.Of.Owner.Occupied.Homes..2018.Est.))
austin_medianincome$No..Of.Households.With.Income..200K...2108.Est. <- as.numeric(gsub(',', '', austin_medianincome$No..Of.Households.With.Income..200K...2108.Est.))
austin_medianincome$No..With.Master.s.Degrees <- as.numeric(gsub(',', '', austin_medianincome$No..With.Master.s.Degrees))
austin_medianincome$No..With.Bachelor.s.Degree <- as.numeric(gsub(',', '', austin_medianincome$No..With.Bachelor.s.Degree))


#filters zip codes only in austin
austin_medianincome = austin_medianincome %>%
  filter(Zip.Code.City. == 'Austin')

#removes column no longer necessary
austin_medianincome = select(austin_medianincome,-3)
```



```{r, echo=FALSE, include=FALSE}
########CLEANS TRAVIS COUNTY PARK DATA
#finds longitude and latitude
travis_parks = travis_parks %>%
  mutate(geocode(FIRST_Address)) 

#finds associated addresses of longitude and latitude
address <- do.call(rbind,
                   lapply(1:nrow(travis_parks),
                          function(i)revgeocode(as.numeric(travis_parks[i,4:5]))))

#combines address vector with austin_parks
travis_parks <- cbind(travis_parks,address)

#extracts zipcode from address vector
travis_parks$zipcode <- substr(str_extract(travis_parks$address," [0-9]{5}, .+"),2,6)

#removes newly created address column 
travis_parks = travis_parks[,-6]

#renames columns in travis parks
travis_parks = rename(travis_parks, address = FIRST_Address)
travis_parks = rename(travis_parks, acres = SUM_Park_Acres)
```


```{r, echo=FALSE, include=FALSE}
##########################CLEANS AUSTIN PARKS DATA##############################
#finds longitude and latitude associated with each park
austin_parks = austin_parks %>%
  mutate(geocode(ADDRESS)) 

#renames columns in Austin parks
austin_parks = rename(austin_parks, address = ADDRESS)
austin_parks = rename(austin_parks, zipcode = ZIPCODE)
austin_parks = rename(austin_parks, acres = ASSET_SIZE)
austin_parks = rename(austin_parks, Name = LOCATION_NAME)

#removes non-Austin zip codes
austin_parks = austin_parks %>%
  filter(zipcode >= 78701)

#removes non-Austin zipcodes
travis_parks = travis_parks %>%
  filter(zipcode >= 78701)

#combines both data sets of austin and travis county parks in austin
all_austin_parks = rbind(austin_parks, travis_parks)
```


```{r, echo=FALSE, include=FALSE}
##########################CLEANS MORE_PARKS DATA################################
#finds longitude and latitude for every address
more_parks = more_parks %>%
  mutate(geocode(Address))

#finds associated addresses of longitude and latitude
address <- do.call(rbind,
                   lapply(1:nrow(more_parks),
                          function(i)revgeocode(as.numeric(more_parks[i,4:5]))))

#combines address vector with austin_parks
more_parks <- cbind(more_parks,address)

#extracts zipcode from address vector
more_parks$zipcode <- substr(str_extract(more_parks$address," [0-9]{5}, .+"),2,6)

#removes newly created address column 
more_parks = more_parks[,-2]

#renames columns in parks
more_parks = rename(more_parks, acres = Acres.of.Land)
more_parks = rename(more_parks, Name = Park.Name)

#combines all park data
all_austin_parks = rbind(all_austin_parks, more_parks)

#reorders columns
column_order <- c("Name", "address", "zipcode", "acres", "lon", "lat")
all_austin_parks = all_austin_parks[,column_order]
```


```{r, echo=FALSE, include=FALSE}
#finds park percentages
zip_totals = summarize(group_by(all_austin_parks, zipcode), total = n(), acres = sum(acres)) %>%
  mutate(share_of_parks = total/sum(total)) %>%
  mutate(avg_acre_in_zip = acres/total)

#adds row for zipcode 78742, which had zero parks
zip_totals = zip_totals %>% 
  add_row(zipcode = '78742', total = 0, acres = 0,share_of_parks = 0, avg_acre_in_zip = 0)

#merges data sets 
merged_data = merge(zip_totals, austin_medianincome)

#parks per capita per 100000
merged_data = merged_data %>%
  mutate(parks_per100000 = total/Population *100000)
```


# IV. Results
```{r, echo=FALSE}
#acerage per 100000
ggplot(data = merged_data, mapping = aes(x = zipcode, y = parks_per100000))+ 
  geom_col(mapping = aes(x = reorder(zipcode, parks_per100000), y = parks_per100000), fill = 'darkgreen') + 
  coord_flip() + scale_y_continuous(name = "Acreage Per 100000") +
  scale_x_discrete(name = "Zipcode") + ggtitle("Figure 1") +
  theme(plot.title = element_text(hjust = 0.5))

```
Figure 1 displays the park distribution in Austin measured in acreage per 100000. Zip code 78701 has the highest acreage per 100000 of 159.12 acres. Alternatively, zip code 78742 the lowest acreage per 100000 of 0 acres. 

```{r, echo=FALSE}
##################HIERARCHICAL CLUSTERING OF MERGED DATA########################
#merges data sets for parks and median income
parks_and_income = merge(all_austin_parks, austin_medianincome)

#removes address column
parks_and_income = select(parks_and_income, -3)

#sums quanities of masters and bachelors degree holders
parks_and_income$education <-rowSums(cbind(parks_and_income$No..With.Master.s.Degrees, parks_and_income$No..With.Bachelor.s.Degree))

#creates new variable of percent of post-secondary education per zip code
parks_and_income= parks_and_income %>%
  mutate(percent_of_education = education/Population)

#creates new interactive variable between income and education
parks_and_income= parks_and_income %>%
  mutate(educ_income = percent_of_education*Median.Household.Income)

```

```{r, echo=FALSE}
X_parks = parks_and_income[,3:16]
X_parks = scale(X_parks, center=TRUE, scale = TRUE)
distances = dist(X_parks, method = "euclidean")
```

```{r, echo=FALSE}
#hierarchical clustering 
cluster_parks = hclust(distances, method = 'ward.D2')

#creates k grid
k_grid = seq(2, 30, by=1)
SSE_grid = foreach(k = k_grid, .combine = 'c') %do% {
  cluster_k = kmeans(X_parks, k, nstart = 50)
  cluster_k$tot.withinss
}
```


```{r, echo=FALSE, include=FALSE, warning=FALSE}
#map of percent of education
p <- ggmap(get_googlemap(center = c(lon =  -97.733330, lat = 30.266666),
                         zoom = 11, scale = 2,
                         maptype ='terrain',
                         color = 'color'))
map1 = p + geom_point(aes(x = lon, y = lat, color = educ_income), 
                         data = parks_and_income, size = 3)+ xlab("Longitude") + ylab("Latitude") + ggtitle("Figure 2") +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r, echo=FALSE}
map1
```
Figure 2 displays a map of Austin and the location of each recorded outdoor space. Moreover, each point displays the result of the interaction variable between post-secondary education as a percentage of population within each zip code and median household income. 

```{r, echo=FALSE}
plot(cluster_parks, labels = FALSE, hang = -1)
```
Using the Ward method to implement hierarchical clustering, each location of an outdoor space is clustered in a well-balanced denodrogram.

```{r, echo=FALSE}
#SSE plot
plot(SSE_grid)
```

Using k-means and the elbow method, k = 2 along the curve is observed to be optimal.


```{r, echo=FALSE}
#cut tree for clusters
clusters <- cutree(cluster_parks, k = 2)

#takes cluster vector and combines with parks_and_income
parks_and_income = cbind(parks_and_income, clusters)

#creates map of clusters
map2 = p + geom_point(aes(x = lon, y = lat, color = factor(clusters)), 
                         data = parks_and_income, size = 3)+ xlab("Longitude") + ylab("Latitude") + ggtitle("Figure 3") +
  theme(plot.title = element_text(hjust = 0.5)) + labs(color = "Cluster")
```

```{r, echo=FALSE}
map2
```

Figure 3 displays a distinguishable pattern of each location of outdoor space in terms of its respective cluster.

```{r, echo=FALSE}
#finds cluster in each park
clust_pop = summarize(group_by(parks_and_income, zipcode), total_parks = n(),
                      Population = sum(Population), clusters = sum(clusters)) %>%
  mutate(Population = Population/total_parks) %>%
  mutate(clusters = clusters/total_parks, clusters = round(clusters, digits = 0))

#sums population in each cluster
clust_pop = summarize(group_by(clust_pop, clusters), Population = sum(Population))

#removes cluster column
clust_pop = select(clust_pop, -1)

#groups clusters and sums parks and acres
clust_totals = summarize(group_by(parks_and_income, clusters), total_parks = n(), Total.Acres = sum(acres)) %>%
  mutate(Share.of.Parks = total_parks/sum(total_parks)) %>%
  mutate(Average.Acres = Total.Acres/total_parks) 

#combines clust_totals and population totals found in clust_pop
clust_totals = cbind(clust_totals, clust_pop)

#calculates acres per 100000
clust_totals = clust_totals%>%
  mutate(Acres.Per.100000 = Total.Acres/Population*100000)

#renames columns
clust_totals = rename(clust_totals, Cluster = clusters)
clust_totals = rename(clust_totals, No.of.Parks = total_parks)

#reorders columns
column_order <- c("Cluster", "Population", "No.of.Parks", "Share.of.Parks", "Total.Acres", "Average.Acres", "Acres.Per.100000")
clust_totals = clust_totals[,column_order]

knitr::kable(clust_totals, "pipe", caption = "Results", digits = 2, col.names = c("Cluster", "Population", "Number of Outdoor Spaces", "Share of Outdoor Spaces", "Total Acres", "Average Acres", "Acres Per 100000"))
```


# V. Conclusion
Outdoor spaces are an important venue for many seeking relaxation, particularly during this time of Covid-19. Austinites have a wide variety of 335 parks spread across 43 zip codes in which to enjoy recreational activity, but equal opportunity to access outdoor spaces may be questionable. Figure 1 displays the distribution of outdoor spaces in terms of acreage per 100000. It becomes obvious upon observation that equal access to outdoor spaces are heavily reliant on the zip code in which you choose to live. By a large difference, zip code 78701 has the highest acreage per 100000 of 159.12 acres. Zip code 78702, the next highest acreage per 100000, had 102.62 acres. Alternatively, zipcode 78742 had the lowest acreage per 100000 of 0 acres. 

When plotting each location and using the interaction between post-secondary education as a percentage of population per zip code and median household income per zip code, there appears to be a slight pattern between East and West Austin, loosely following I-35. Hierarchical clustering was used to assist in better understanding the proposed pattern. Once choosing an optimal k clusters of 2, plotting the results became more distinguishable, as noted in Table 1. 

Table 1 displays each clusters respective population, number of parks, percentage share of parks, total acres, average acre, and acres per 100000. Cluster 1 not only has a smaller total number and share of parks, but also a smaller population. However, cluster 1 has a larger amount of total acres, average acres, and acres per 100000. Relative to cluster 2, cluster 1 has 22.1% smaller population, but 30.5% more total acres, 39.4% higher average acre, and 67.5% more acres per 100000. In conjunction with Figure 1 and Figure 2, there appears to be a correlation with post-secondary education and wealth with regards to access to outdoor spaces in Austin, Texas. 









